\chapter[Standards \& Constraints]{Standards and Design Constraints}
\label{ch:standards_and_design_constraints}

The development and evaluation of the passive anxiety detection system involved strict adherence to international engineering standards, ethical data management principles, and multiple practical design constraints. Because the system operates at the intersection of supervised machine learning, OS-level behavioral telemetry, and human psychological measurement, its design choices must be evaluated not only against conventional software engineering benchmarks but also against the complex human-centered and ethical criteria that govern affective computing research. This chapter presents a complete compliance mapping, design constraint analysis, and complex engineering problem verification for the project.

\section{Compliance with Standards}
\label{sec:standards}

\subsection{Software Engineering Standards}
The telemetry collection pipeline and machine learning data processing components are aligned with the following foundational software engineering standards:

\begin{itemize}
    \item \textbf{IEEE 830-1998 (Software Requirements Specifications):} The functional and non-functional requirements of the background monitoring script---including the 30-second window aggregation cycle, the 18-feature output schema, and the silent \texttt{.bat} launcher behavior---were formally specified in accordance with this standard \cite{ieee830}. Compliance ensures deterministic, traceable operation and prevents race conditions from affecting the primary Code::Blocks development environment.

    \item \textbf{IEEE 7000-2021 (Ethical Concerns During System Design):} Because the system processes sensitive cognitive and affective behavioral metadata from real human participants, this standard was adopted as the governing framework for ethical data handling \cite{ieee7000}. All telemetry extraction boundaries are explicitly defined: no literal keystrokes, source code characters, or alphanumeric input sequences are ever recorded. Only timing intervals, event counts, and window classification labels are persisted, ensuring full transparency and minimizing privacy exposure.

    \item \textbf{ISO/IEC 27001 (Information Security Management Systems):} All student behavioral data is stored exclusively in local CSV files on the individual student workstation \cite{isoiec27001}. No data is transmitted over any network channel during or after the monitoring session. Post-session data collection is performed manually by the researcher via physical file transfer, maintaining complete isolation from external access threats and ensuring compliance with institutional data security obligations.
\end{itemize}

\subsection{Human-System Interaction Standards}
The passive monitoring architecture is designed in alignment with established human-system interaction principles:

\begin{itemize}
    \item \textbf{ISO 9241-210:2019 (Human-Centred Design for Interactive Systems):} This standard requires that monitoring tools minimize user cognitive burden, preserve task continuity, and remain fully transparent in their operation \cite{iso9241}. The passive monitoring system satisfies all three requirements: the \texttt{.bat} launcher operates with zero visible interface, the background hook adds no measurable latency to the Code::Blocks compilation pipeline, and no interaction from the student is ever required or expected during data collection.

    \item \textbf{Windows API Compatibility (WinAPI):} The low-level keyboard hook and window focus tracking components utilize native Windows API calls (\texttt{SetWindowsHookEx}, \texttt{GetForegroundWindow}), ensuring full compatibility with standard Windows 10/11 university workstation configurations without requiring elevated administrator privileges or custom driver installation.
\end{itemize}

\subsection{Data Handling and Ethical Research Standards}
\begin{itemize}
    \item \textbf{Informed Consent Protocol:} All 67 student participants were briefed by the supervising research team on the purpose of the data collection prior to the commencement of each examination session. Participation was voluntary, and all data was anonymized at the point of collection using session-level identifiers rather than personal names or student ID numbers.

    \item \textbf{Post-Exam Survey Ethics:} The Likert-scale stress survey was administered within 5--15 minutes of exam completion to maximize psychological recall accuracy. Students were clearly informed that survey responses would not affect their academic grades in any way, eliminating social-desirability bias in self-reporting.
\end{itemize}

\section{Design Constraints}
\label{sec:constraints}

\subsection{Economic Constraint}
Conventional psychological informatics and physiological stress-monitoring systems depend on expensive biometric hardware, including multi-channel EEG caps (typically \$3,000--\$15,000 per unit), clinical Galvanic Skin Response (GSR) wristbands, and high-resolution eye-tracking rigs \cite{mdpi389, core579951358}. The financial overhead of such equipment creates an insurmountable structural barrier for academic laboratories operating under standard university budget allocations. This project was developed and deployed under a strict zero-capital constraint: the entire data collection and analysis pipeline was built exclusively on software components, utilizing the standard student keyboards, workstations, and operating systems already present in the university computing lab.

\subsection{Environmental Constraint}
The system is designed to operate exclusively within the Code::Blocks C/C++ IDE environment on Windows workstations, as this reflects the specific academic context in which the data was collected \cite{acm4897123}. The deployment environment is inherently unregulated: student behaviors vary widely, exam room temperatures, seating configurations, and keyboard hardware differ between sessions, and background noise from other students is present. The monitoring system must remain robust and consistent under all these variable real-world conditions without requiring any standardized physical setup or controlled lab configuration \cite{researchgateKeystroke}.

\subsection{Ethical Constraint}
The management of user behavioral interaction telemetry data creates immediate privacy obligations that must be addressed at the architectural level. The background data extraction hook is specifically designed to exclude all literal text characters, alphanumeric keystroke sequences, and code content from its logging arrays. The system operates exclusively on structural behavioral patterns—tracking non-literal variables such as inter-keystroke timing intervals, backspace event counts, window focus transition timestamps, and compilation event frequencies. This data boundary ensures that no proprietary source code, personal credentials, or sensitive communication content can be captured, stored, or exposed, maintaining full compliance with institutional ethical data protection standards.

\subsection{Health and Safety Constraint}
Because the monitoring system operates on real human participants under exam stress conditions, specific care was taken to ensure that the system itself does not exacerbate participant anxiety. The complete invisibility of the monitoring script to the student eliminates the ``observer effect''---the phenomenon where participants alter their natural behavior upon awareness of being monitored. All student interaction data was de-identified immediately upon collection, and individual-level data was accessible only to the core research team under a defined data access protocol approved by the institutional faculty supervisor.

\subsection{Social Constraint}
The computer science academic community and software engineering industry face a significant shortfall in forward-looking mental health support infrastructure, contributing to elevated rates of early-career burnout, imposter syndrome, and chronic workplace stress among programmers \cite{mdpi5739, codereview2024, stemedu2025}. This project responds to this social constraint by developing a scalable, software-only behavioral analytics system capable of passively flagging student distress without requiring physical intervention, specialized clinical staff, or dedicated mental health hardware \cite{selfreg2023}. The findings validate that passive behavioral monitoring can serve as a non-invasive early warning mechanism embedded within existing educational infrastructure.

\subsection{Sustainability and Maintenance Constraint}
The machine learning classification pipeline is built entirely on open-source Python libraries (Scikit-learn, Pandas, Imbalanced-learn), ensuring long-term maintainability without dependency on commercial software licenses or subscription services \cite{rahimi2024ai}. The modular architecture of the pipeline---with clearly separated data merging, preprocessing, SMOTE balancing, and model evaluation stages---enables any future researcher to independently update individual components (e.g., replacing the SVM classifier with a deep learning model) without rebuilding the entire system from scratch \cite{arean2025generalizable}.

\section{Cost Analysis and Feasibility Model}
\label{sec:cost}

The passive anxiety detection system was developed and deployed at a total direct financial cost of \textbf{\$0.00 (zero taka)}. Table~\ref{tab:cost_analysis} provides an itemized breakdown demonstrating how each component of the system avoided commercial expenditure.

\begin{table}[htbp]
    \centering
    \small
    \caption{Itemized Cost Analysis of the Passive Anxiety Detection System.}
    \label{tab:cost_analysis}
    \begin{tabular}{|p{4.5cm}|p{4.5cm}|p{3.5cm}|}
        \hline
        \textbf{System Component} & \textbf{Implementation Approach} & \textbf{Cost} \\ \hline
        Data Collection Hardware & Standard university lab keyboards and workstations (pre-existing) & \$0.00 \\ \hline
        Monitoring Script Runtime & Python 3 (open-source, MIT license) & \$0.00 \\ \hline
        OS Keyboard Hook & Native Windows API (WinAPI) — no SDK required & \$0.00 \\ \hline
        Data Storage & Local CSV files on student workstations & \$0.00 \\ \hline
        ML Pipeline & Scikit-learn, Pandas, Imbalanced-learn (open-source) & \$0.00 \\ \hline
        Statistical Analysis & SciPy, Matplotlib (open-source) & \$0.00 \\ \hline
        Network Infrastructure & None required — fully offline system & \$0.00 \\ \hline
        \textbf{Total Project Cost} & & \textbf{\$0.00} \\ \hline
    \end{tabular}
\end{table}

\noindent This zero-cost architecture is a key contribution of this research. By demonstrating that a statistically validated, machine-learning-based anxiety detection system can be built entirely without hardware procurement or commercial software licenses, this project establishes a replicable and financially accessible blueprint for universities worldwide, including institutions in low-resource settings where EEG or biometric monitoring equipment is entirely out of reach.

\section[Problem Mapping]{Complex Engineering Problem Mapping}
\label{sec:cep}

The development and evaluation of this passive developer tracking system qualifies as a \textbf{Complex Engineering Problem} under international engineering accreditation rubrics, as it integrates advanced machine learning classification, low-level OS hooks, statistical signal validation, and human-centered research ethics to resolve an abstract, psychologically grounded challenge.

\subsection{Complex Problem Solving (P1--P7)}
Table~\ref{tab:p_solve} presents the verified mapping of the project's characteristics against the seven recognized Complex Engineering Problem criteria (P1--P7).

\begin{table}[htbp]
    \centering
    \small
    \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        \textbf{P1} & \textbf{P2} & \textbf{P3} & \textbf{P4} & \textbf{P5} & \textbf{P6} & \textbf{P7} \\ \hline
        \begin{tabular}{@{}c@{}}Depth of \\ Knowledge\end{tabular} &
        \begin{tabular}{@{}c@{}}Conflicting \\ Requirements\end{tabular} &
        \begin{tabular}{@{}c@{}}Depth of \\ Analysis\end{tabular} &
        \begin{tabular}{@{}c@{}}Familiarity \\ of Issues\end{tabular} &
        \begin{tabular}{@{}c@{}}Applicable \\ Standards\end{tabular} &
        \begin{tabular}{@{}c@{}}Stakeholder \\ Involvement\end{tabular} &
        \begin{tabular}{@{}c@{}}Inter- \\ dependence\end{tabular} \\ \hline
        $\checkmark$ & $\checkmark$ & $\checkmark$ & $\times$ & $\checkmark$ & $\checkmark$ & $\checkmark$ \\ \hline
    \end{tabular}
    \caption{Mapping Project Attributes to Complex Engineering Problem Solving Criteria (P1--P7).}
    \label{tab:p_solve}
\end{table}

\begin{itemize}
    \item \textbf{P1 — Depth of Knowledge:} The project requires advanced multi-domain engineering knowledge spanning supervised machine learning (SVM hyperplane optimization, SMOTE interpolation), operating system API programming (asynchronous WinAPI keyboard hooks), statistical signal validation (Pearson and Spearman correlation), and applied cognitive psychology (mapping inter-keystroke timing variance to psychological arousal models). No single engineering discipline provides sufficient knowledge to address all aspects of the problem.

    \item \textbf{P2 — Range of Conflicting Requirements:} The system architecture must simultaneously satisfy fundamentally competing engineering objectives. Maximum classification accuracy demands high-frequency, richly detailed behavioral feature extraction across all application window contexts. However, strict user privacy protection requires that no literal character data or identifiable content is ever recorded. Simultaneously, minimal CPU overhead must be maintained to ensure that the monitoring process never delays compiler execution in the primary Code::Blocks workspace. Resolving these three conflicting constraints required careful asynchronous thread design and a precisely scoped feature extraction boundary.

    \item \textbf{P3 — Depth of Analysis:} Human anxiety is an inherently non-linear, context-dependent, and subjectively experienced psychological state. It cannot be reliably captured through simple threshold rules or basic linear classifiers. The project demands deep mathematical analysis, including the construction of a rule-based anxiety scoring engine with weighted feature normalization, application of the SMOTE interpolation formula for minority class synthesis, and evaluation of eight classifiers across 100 independent cross-validation runs to produce statistically reliable, variance-quantified performance estimates.

    \item \textbf{P5 — Extent of Applicable Standards:} The design process operates under multiple overlapping international standards simultaneously. IEEE 7000-2021 governs the ethical boundaries of affective data extraction. ISO/IEC 27001 constrains the data storage and access control architecture. ISO 9241-210:2019 defines the non-intrusiveness and transparency requirements for the human-system interaction. Navigating these standards concurrently while maintaining technical performance constitutes a non-trivial multi-constraint engineering problem.

    \item \textbf{P6 — Stakeholder Involvement:} The system directly involves three distinct stakeholder groups with different and partially conflicting interests: (1) first-year computer science students who are the data subjects and whose cognitive experience must not be disrupted; (2) university faculty and exam invigilators who oversee the deployment environment and whose cooperation is required for ethical approval; and (3) the academic research committee who demand rigorous statistical validation, reproducible results, and compliance with thesis documentation standards.

    \item \textbf{P7 — Inter-dependence:} The project sub-systems exhibit tight causal dependencies. If the \texttt{.bat} launcher fails to initialize on a student workstation, no CSV log is produced for that session, directly reducing the dataset size and degrading ML model generalizability. If the data merging pipeline introduces a labeling error during post-exam CSV consolidation, the Likert survey ground truth labels are misaligned, and all downstream classification results become invalid. Each sub-system failure propagates irreversibly through the full research pipeline.
\end{itemize}

\subsection{Complex Engineering Activities (A1--A5)}
Table~\ref{tab:e_act} documents the mapping of research activities against the five international criteria for Complex Engineering Activities (A1--A5).

\begin{table}[htbp]
    \centering
    \small
    \begin{tabular}{|c|c|c|c|c|}
        \hline
        \textbf{A1} & \textbf{A2} & \textbf{A3} & \textbf{A4} & \textbf{A5} \\ \hline
        \begin{tabular}{@{}c@{}}Range of \\ Resources\end{tabular} &
        \begin{tabular}{@{}c@{}}Level of \\ Interaction\end{tabular} &
        \begin{tabular}{@{}c@{}}Innovation\end{tabular} &
        \begin{tabular}{@{}c@{}}Consequences for \\ Society \& Environment\end{tabular} &
        \begin{tabular}{@{}c@{}}Familiarity\end{tabular} \\ \hline
        $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\times$ \\ \hline
    \end{tabular}
    \caption{Mapping Research Activities to Complex Engineering Activity Criteria (A1--A5).}
    \label{tab:e_act}
\end{table}

\begin{itemize}
    \item \textbf{A1 — Range of Resources:} The research required the coordinated utilization of diverse technical and human resources: a Python-based machine learning pipeline, a Windows API behavioral telemetry engine, a cohort of 67 live human research participants, four separate graded examination sessions coordinated with university faculty, and post-exam Likert survey administration logistics. Managing these heterogeneous resources simultaneously under academic time constraints constitutes a complex engineering activity.

    \item \textbf{A2 — Level of Interaction:} The project required sustained interdisciplinary collaboration across OS-level systems programming (the \texttt{.bat} hook development), data engineering (CSV merging and outlier conditioning), statistical analysis (Pearson/Spearman correlation validation), and machine learning (SMOTE + 10$\times$10 cross-validation pipeline). Each sub-discipline informed the design constraints of the others, requiring continuous iterative coordination across the full research team.

    \item \textbf{A3 — Innovation:} The system introduces a novel paradigm for developer well-being instrumentation. Unlike prior approaches that depend on physiological wearable hardware or subject-dependent personalized calibration, this framework delivers a fully software-based, subject-independent universal classifier trained on real exam-stress behavioral data from a live cohort. The combination of a rule-based heuristic scoring engine validated against a 10$\times$10 leakage-free cross-validation ML pipeline represents a methodologically original contribution to the HCI and educational analytics fields.

    \item \textbf{A4 — Consequences for Society and Environment:} The deployment of this system has direct positive societal implications. It provides academic institutions with a zero-cost mechanism to passively identify student programmers who are experiencing examination anxiety at clinically significant levels, enabling early pedagogical intervention before performance degradation or course dropout occurs. Environmentally, the entirely software-based, zero-hardware architecture eliminates e-waste generation and toxic battery disposal concerns that accompany biometric wearable device deployments at scale.
\end{itemize}

\section{Summary}
\label{sec:ch5_summary}
This chapter established the complete compliance and constraint framework governing the passive anxiety detection system. The architecture aligns with three international standards—IEEE 830-1998, IEEE 7000-2021, and ISO/IEC 27001—while being explicitly designed to satisfy the conflicting engineering constraints of maximal privacy protection, zero CPU overhead, and deployment-ready ecological validity within real university examination environments. The zero-cost itemized feasibility model confirms the financial accessibility of the system at institutional scale. The verified mappings against Complex Engineering Problem criteria P1--P7 and Complex Engineering Activity criteria A1--A5 formally establish the academic and technical rigor of this research within international engineering accreditation standards.